Types for R, Empirically Designed and Evaluated

### 2020-05-07 paper/main.tex
Itemize contractr challenges

%
%
\subsubsection{Challenges}
Retrofitting a robust and efficient type-system as an external package in R has
been a significant undertaking. We had to contend with the complex design
choices of R to make it work without surprises for the end-users. We describe
two key challenges that we faced in the development of \contractr.
Firstly, we faced the problem of type-checking arguments in a non-strict
language while retaining the non-strict semantics. When a function is called in
R, parameters are bound to unevaluated code thunks called
promises~\cite{oopsla19}, instead of values which can be immediately
type-checked. Naively forcing promises at function call to obtain a value for
type-checking would violate the language semantics, leading to incorrect
results. To preserve the original semantics, \contractr modifies the
promise by wrapping the unevaluated promise expression in a call to its
type-checker along with the context necessary to associate the promise to the
function and its parameter (for failure messages). The contract checking happens
when the promise is evaluated, either inside the function or many calls deep.
This works, but with one wrinkle. GNU R has a bytecode compiler that can
sometimes optimize away promises; in those cases, \contractr receives
values that can be immediately type-checked. Non-strict semantics dictate that
any computation related to the argument should happen on the first use of the
argument. So type-checking at this point and issuing an error message would
violate the non-strict semantics; if this argument is never used, it would be
incorrect to type-check it at the beginning of call. To make this work,
\contractr mutates the function parameter binding to a promise that wraps
the argument value in a call to the type-checker, as in the previous case.
Secondly, we faced the problem of type-checking function return value. In R, the
result of a function call is the result of evaluating the last expression in the
function body; so an explicit return call is absent from most function
definitions. Thus, there could be many potential sub-expressions in the function
body which need to be type-checked aginst the return type. To address this,
\contractr registers the type-checker to be called on the return value
through a function exit hook. This hook is executed in the function call's
environment after it has executed. While this works, we have to contend with yet
another wrinkle, R interpreter can perform longjumps in its C implementation,
which causes active R function calls on the stack to be discarded. When they are
discarded, their exit hooks are called, which in this case calls the registered
type-checker. But, these functions are in the middle of an active computation,
so they don't have a return value to type-check. \contractr deals with
this problem by allocating a unique sentinel object which serves as the return
value for calls that are discarded. The function call exit hook does not call
the type-checker if this unique value happens to be the return value of the
call.

%
%
\subsubsection{Usability}
With \contractr, we aim to provide a hassle-free path to the R developers for
adding types to untyped package code. Hence, we have paid a lot of attention
towards usability in its design. We discuss three such usability features.
\begin{itemize}

>>>>

%
%
\subsubsection{Challenges}
Retrofitting a robust and efficient type-system as an external package in R has
been a significant undertaking. We had to contend with the complex design
choices of R to make it work without surprises for the end-users. We describe
two key challenges that we faced in the development of \contractr.
\begin{itemize}
\item Firstly, we faced the problem of type-checking arguments in a non-strict
  language while retaining the non-strict semantics. When a function is called in
  R, parameters are bound to unevaluated code thunks called
  promises~\cite{oopsla19}, instead of values which can be immediately
  type-checked. Naively forcing promises at function call to obtain a value for
  type-checking would violate the language semantics, leading to incorrect
  results. To preserve the original semantics, \contractr modifies the
  promise by wrapping the unevaluated promise expression in a call to its
  type-checker along with the context necessary to associate the promise to the
  function and its parameter (for failure messages). The contract checking happens
  when the promise is evaluated, either inside the function or many calls deep.
  This works, but with one wrinkle. GNU R has a bytecode compiler that can
  sometimes optimize away promises; in those cases, \contractr receives
  values that can be immediately type-checked. Non-strict semantics dictate that
  any computation related to the argument should happen on the first use of the
  argument. So type-checking at this point and issuing an error message would
  violate the non-strict semantics; if this argument is never used, it would be
  incorrect to type-check it at the beginning of call. To make this work,
  \contractr mutates the function parameter binding to a promise that wraps
  the argument value in a call to the type-checker, as in the previous case.
\item Secondly, we faced the problem of type-checking function return value. In R, the
  result of a function call is the result of evaluating the last expression in the
  function body; so an explicit return call is absent from most function
  definitions. Thus, there could be many potential sub-expressions in the function
  body which need to be type-checked aginst the return type. To address this,
  \contractr registers the type-checker to be called on the return value
  through a function exit hook. This hook is executed in the function call's
  environment after it has executed. While this works, we have to contend with yet
  another wrinkle, R interpreter can perform longjumps in its C implementation,
  which causes active R function calls on the stack to be discarded. When they are
  discarded, their exit hooks are called, which in this case calls the registered
  type-checker. But, these functions are in the middle of an active computation,
  so they don't have a return value to type-check. \contractr deals with
  this problem by allocating a unique sentinel object which serves as the return
  value for calls that are discarded. The function call exit hook does not call
  the type-checker if this unique value happens to be the return value of the
  call.
\end{itemize}
%
%
\subsubsection{Usability}
With \contractr, we aim to provide a hassle-free path to the R developers for
adding types to untyped package code. Hence, we have paid a lot of attention
towards usability in its design. We discuss three such usability features.
\begin{itemize}
